---
title: "CEE 218X Assignment 1"
author: "Darren Wong"
date: 9/25/2021
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```

```{r set directory}
setwd("~/Documents/GitHub/darrenwong.github.io")
library(tidyverse)
```

```{r data input for electricity, include=FALSE}
year <- 2017
quarters <- 1:4
type <- "Electric"

pge_17_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_17_elec <- rbind(pge_17_elec,temp)
  saveRDS(pge_17_elec, "pge_17_elec.rds")
}


year <- 2018
pge_18_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_18_elec <- rbind(pge_18_elec,temp)
  saveRDS(pge_18_elec, "pge_18_elec.rds")
}

year <- 2019
pge_19_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_19_elec <- rbind(pge_19_elec,temp)
  saveRDS(pge_19_elec, "pge_19_elec.rds")
}

year <- 2020
pge_20_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_20_elec <- rbind(pge_20_elec,temp)
  saveRDS(pge_20_elec, "pge_20_elec.rds")
}

year <- 2021
quarters <- 1:2
pge_21_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_21_elec <- rbind(pge_21_elec,temp)
  saveRDS(pge_21_elec, "pge_21_elec.rds")
}

pge_17_21_elec <- do.call("rbind", list(pge_17_elec, pge_18_elec, pge_19_elec, pge_20_elec, pge_21_elec))

saveRDS(pge_17_21_elec, "pge_17_21_elec.rds")
```

```{r data input for gas, include=FALSE}
year <- 2017
quarters <- 1:4
type <- "Gas"

pge_17_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_17_gas <- rbind(pge_17_gas,temp)
  saveRDS(pge_17_gas, "pge_17_gas.rds")
}

year <- 2018
pge_18_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_18_gas <- rbind(pge_18_gas,temp)
  saveRDS(pge_18_gas, "pge_18_gas.rds")
}

year <- 2019
pge_19_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_19_gas <- rbind(pge_19_gas,temp)
  saveRDS(pge_19_gas, "pge_19_gas.rds")
}

year <- 2020
pge_20_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_20_gas <- rbind(pge_20_gas,temp)
  saveRDS(pge_20_gas, "pge_20_gas.rds")
}

year <- 2021
quarters <- 1:2
pge_21_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_21_gas <- rbind(pge_21_gas,temp)
  saveRDS(pge_21_gas, "pge_21_gas.rds")
}

pge_17_21_gas <- do.call("rbind", list(pge_17_gas, pge_18_gas, pge_19_gas, pge_20_gas, pge_21_gas))

saveRDS(pge_17_21_gas, "pge_17_21_gas.rds")
```

```{r merge datafiles}
pge_final <- merge (x=pge_17_21_elec, y=pge_17_21_gas, all=TRUE)
pge_filter_res <- filter(pge_final, CUSTOMERCLASS %in% c("Elec- Residential", "Gas- Residential"))
pge_filter_com <- filter(pge_final, CUSTOMERCLASS %in% c("Elec- Commercial", "Gas- Commercial"))
```

```{r data preprocessing}
pge_res_final <-
  pge_filter_res %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_res_final2 <-
  pge_filter_res %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_com_final <-
  pge_filter_com %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU; no conversion needed for gas as 1therm = 1kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_com_final2 <-
  pge_filter_com %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU; no conversion needed for gas as 1therm = 1kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )
```

# PG&E Monthly Electricity and Gas Usage (Residential)

The total electricity and gas usage fluctuates seasonally, peaking each winter between November and March at about 3.5 to 4.0 x 10^10^ kBTU. Generally, the seasonal fluctuation in total energy use is driven by gas usage as electricity usage remains relatively consistent each month, and this can be explained by the role of heating in building systems during winter.

Between 2019 and 2021, there has been higher peaks in total energy use from about 3.5 x 10^10^ kBTU in 2019 to almost 4 x 10^10^ kBTU in 2021. The increase in energy consumption could be ascribed to people spending more time in their homes for work and education during the COVID-19 pandemic, thereby requiring more energy for building heating.

There appears to be an anomalously high electricity and gas usage in September 2017, which may be due to a flawed dataset.

```{r electricity and gas residential chart - date}
library(plotly)

pge_chart_res1 <- 
  pge_res_final %>%
  ggplot() +
  geom_bar(
    aes(
      x = DATE %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Residential) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_res1 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)
```

A key issue about analysing large datasets (i.e. 54 months across many zipcodes) is that it becomes hard to observe cumulative trends, e.g. by year rather than month. Grouping the data of residential sector into years (2021 data only includes the first two quarters), there is a clear increase in total energy usage between 2019 and 2020 during the COVID-19 pandemic. The interactive plot allows us to identify a bigger jump in electricity than gas usage (in absolute terms). Perhaps gas usage is less dynamic than electricity usage as heating is likely the main source and is less variable, compared to the diverse range of household and electrical applicances that require electricity when people are at home.

```{r electricity and gas residential chart - year}
pge_chart_res2 <- 
  pge_res_final2 %>%
  ggplot() +
  geom_bar(
    aes(
      x = YEAR %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Residential) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_res2 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```

# PG&E Monthly Electricity and Gas Usage (Commercial)

On the other hand, total electricity and gas usage in the commercial sector are more balanced. There is also a seasonal fluctuation of greater energy use over winter, likely due to heating. There appears to be a sharp decrease in total energy use between February and April 2020 from 1.5 to almost 1.0 x 10^10^ kBTU when the first wave of the COVID-19 pandemic occurred, which can be attributed to the introduction of remote working to reduce social interaction. However, more recent data suggests that people are returning to offices, with total energy use in the first half of 2021 showing a smaller decline compared to 2020.

The anomalously high electricity and gas usage in September 2017 is also present in this dataset.


```{r electricity and gas commercial chart}
pge_chart_com1 <- 
  pge_com_final %>%
  ggplot() +
  geom_bar(
    aes(
      x = DATE %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Commercial) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))

pge_chart_com1 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```

Likewise, grouping the data of the commercial sector into years (2021 data only includes the first two quarters), total energy consumption in 2020 was lower than that in 2019. This is expected due to more people working outside offices. However, the data also shows fluctuations in total energy use from 2017 to 2020, rather than an upward trend that was 'broken' by COVID-19. It must be acknowledged that other factors beyond the COVID-19 pandemic influencing energy use are also at play, such as cost of energy. While we may anticipate an increasing trend of energy consumption over time due to affluence, it may also be mitigated by better design of urban systems to improve energy efficience.  

```{r electricity and gas commercial chart - year}
pge_chart_com2 <- 
  pge_com_final2 %>%
  ggplot() +
  geom_bar(
    aes(
      x = YEAR %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Commercial) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_com2 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)
```


