---
title: "CEE 218X Assignment 3"
author: "Darren Wong"
date: 10/08/2021
output: html_document
---

```{r setup, include=FALSE, warning=F, message = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r, include = FALSE}
library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r, include = FALSE}
temp <- tempfile()
download.file("https://www2.census.gov/programs-surveys/acs/data/pums/2019/1-Year/csv_hca.zip",destfile = temp)

pums_hca_2019_1yr <- read_csv(unzip(temp,"psam_h06.csv"))

unlink(temp)
```

```{r, include = FALSE}
pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "NP",
    "HHL",
    "HINCP",
    "WATP",
    "WATFP",
    "WAGP",
    "AGEP"
  )
)
```

```{r}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

bay_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% bay_pumas$PUMACE10)
```

```{r}
bay_pums_x1 <-
  bay_pums %>% 
  mutate(
    PWGTP = as.numeric(PWGTP),
    old_watercost_lowinc = ifelse(
        (AGEP >= 60) &
        (WATFP == 3) &
        (WATP > 2000) &
        (WAGP < 30000),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    sum_old_watercost_lowinc = 
      sum(old_watercost_lowinc, na.rm =T),
    perc_old_watercost_lowinc =
      sum(old_watercost_lowinc, na.rm =T)/sum(PWGTP, na.rm = T)*100
  ) %>% 
  left_join(
    bay_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()
```

```{r}
sum <- sum(bay_pums_x1$sum_old_watercost_lowinc)
```


## Vulnerability of Old Individuals Aged 60+ to High Water Prices during Heat Waves in the Bay Area

```{r}
pums_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_pums_x1$perc_old_watercost_lowinc
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = bay_pums_x1,
    fillColor = ~pums_pal(perc_old_watercost_lowinc),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(perc_old_watercost_lowinc), 
      "% individuals aged 60+ making less than $30K and paying more than $2K/year for water"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = bay_pums_x1,
    pal = pums_pal,
    values = ~perc_old_watercost_lowinc,
    title = "% individuals aged<br>60+ making less<br>than $30K and paying<br>more than $2K/yr for<br>water"
  )
```

The criteria (and PUMS indicators) selected in this analysis are older individuals aged 60 and above (AGEP) who make less than $30,000 a year (WAGP) and pay more than $2,000 a year for water (WATP). Individuals are filtered out if they have valid annual water costs (WATFP), rather than being included in their rent or free of charge. Heatwaves are associated with increased mortality in older people due to their aging body's decreased capacity to adapt to changes in body temperature. Low-income elderly are less able to take measures to protect themselves, such as through cooling units, and this problem is exacerbated by the cost of securing water to stay cool and hydrated. The price of water has increased at six times the rate of inflation across California. Water prices are increasing in California as water is often obtained from distant sources for several reasons: arid climate, large infrastructure projects and increasing frequency of heat waves, droughts and hot spells. 

The map shows the percentage of older individuals aged 60 and above making less than \$30,000 a year and paying more than \$2,000 year for water in the Bay Area. There are several parts of the Bay Area with a significant percentage of vulnerable individuals, including Marin County (17%), North Richmond (14%), Moraga (15%) and Los Gatos (12%). In total, 675596 individuals are vulnerable. During heat waves, water prices further increase due to demand-supply mechanisms, exacerbating old individuals' health vulnerability. Policy interventions are therefore needed to ensure access to safe, clean and affordable water, such as rate assistance for these individuals and more flexible payment options to reduce service cuts during heatwaves. 





