---
title: "CEE 218X Assignment 3"
author: "Darren Wong"
date: 10/08/2021
output: html_document
---

```{r setup, include=FALSE, warning=F, message = F}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

```{r, include = FALSE}
library(censusapi)
library(tidyverse)
library(tigris)
library(sf)
library(leaflet)

Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r, include = FALSE}
temp <- tempfile()
download.file("https://www2.census.gov/programs-surveys/acs/data/pums/2019/1-Year/csv_hca.zip",destfile = temp)

pums_hca_2019_1yr <- read_csv(unzip(temp,"psam_h06.csv"))

unlink(temp)
```

```{r, include = FALSE}
pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "NP",
    "HHL",
    "HINCP",
    "WATP",
    "WATFP"
  )
)
```

```{r}
ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

bay_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% bay_pumas$PUMACE10)
```

```{r}
bay_pums_x1 <-
  bay_pums %>% 
  filter(!duplicated(SERIALNO)) %>% 
  mutate(
    WGTP = as.numeric(WGTP),
    nonenglish_watercost_lowinc = ifelse(
      (HHL != 1) &
        (WATFP == 3) &
        (WATP > 2000) &
        (HINCP < 80000),
      WGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    sum_nonenglish_watercost_lowinc = 
      sum(nonenglish_watercost_lowinc, na.rm =T),
    perc_nonenglish_watercost_lowinc =
      sum(nonenglish_watercost_lowinc, na.rm =T)/sum(WGTP, na.rm = T)*100
  ) %>% 
  left_join(
    bay_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()
```

```{r}
sum <- sum(bay_pums_x1$sum_nonenglish_watercost_lowinc)
```


## Vulnerability of Low-Income, Non-English-Speaking Households to High Water Prices during Heat Waves in the Bay Area

```{r}
pums_pal <- colorNumeric(
  palette = "Oranges",
  domain = bay_pums_x1$perc_nonenglish_watercost_lowinc
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = bay_pums_x1,
    fillColor = ~pums_pal(perc_nonenglish_watercost_lowinc),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(perc_nonenglish_watercost_lowinc), 
      "% non-English-speaking households making less than $80K and paying more than $2K/year for water"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = bay_pums_x1,
    pal = pums_pal,
    values = ~perc_nonenglish_watercost_lowinc,
    title = "% non-English-speaking<br>households making less<br>than $80K and paying<br>more than $2K/yr for<br>water"
  )
```

The price of water has increased at six times the rate of inflation across California. Water prices are increasing in California as water is often obtained from distant sources for several reasons: arid climate, large infrastructure projects and increasing frequency of heat waves, droughts and hot spells. The map shows the percentage of non-English-speaking households making less than \$80,000 and paying more than \$2,000 year for water in the Bay Area. There are several parts of the Bay Area with a significant percentage of vulnerable households, including South San Francisco (more than 20%), Fremont (30 to 33%) and northern and eastern San Jose (30%). In total, 409485 households are vulnerable. During heat waves, water prices further increase due to demand-supply mechanisms. Policy interventions are needed to ensure access to safe, clean and affordable water, such as rate assistance for these households and more flexible payment options to reduce service cuts during heatwaves. 





