---
title: "CEE 218X Assignment 1"
author: "Darren Wong"
date: 9/25/2021
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = F, warning = F, echo = F)
```

```{r set directory}
setwd("~/Documents/GitHub/darrenwong.github.io")
library(tidyverse)
```

```{r data input for electricity, include=FALSE}
year <- 2017
quarters <- 1:4
type <- "Electric"

pge_17_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_17_elec <- rbind(pge_17_elec,temp)
  saveRDS(pge_17_elec, "pge_17_elec.rds")
}

pge_17_elec <- pge_17_elec[!duplicated(pge_17_elec[c('ZIPCODE','YEAR', 'MONTH', 'CUSTOMERCLASS')]),]

year <- 2018
pge_18_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_18_elec <- rbind(pge_18_elec,temp)
  saveRDS(pge_18_elec, "pge_18_elec.rds")
}

year <- 2019
pge_19_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_19_elec <- rbind(pge_19_elec,temp)
  saveRDS(pge_19_elec, "pge_19_elec.rds")
}

year <- 2020
pge_20_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_20_elec <- rbind(pge_20_elec,temp)
  saveRDS(pge_20_elec, "pge_20_elec.rds")
}

year <- 2021
quarters <- 1:2
pge_21_elec <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_21_elec <- rbind(pge_21_elec,temp)
  saveRDS(pge_21_elec, "pge_21_elec.rds")
}

pge_17_21_elec <- do.call("rbind", list(pge_17_elec, pge_18_elec, pge_19_elec, pge_20_elec, pge_21_elec))

saveRDS(pge_17_21_elec, "pge_17_21_elec.rds")
```

```{r data input for gas, include=FALSE}
year <- 2017
quarters <- 1:4
type <- "Gas"

pge_17_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_17_gas <- rbind(pge_17_gas,temp)
  saveRDS(pge_17_gas, "pge_17_gas.rds")
}

pge_17_gas <- pge_17_gas[!duplicated(pge_17_gas[c('ZIPCODE','YEAR', 'MONTH', 'CUSTOMERCLASS')]),]

year <- 2018
pge_18_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_18_gas <- rbind(pge_18_gas,temp)
  saveRDS(pge_18_gas, "pge_18_gas.rds")
}

year <- 2019
pge_19_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_19_gas <- rbind(pge_19_gas,temp)
  saveRDS(pge_19_gas, "pge_19_gas.rds")
}

year <- 2020
pge_20_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_20_gas <- rbind(pge_20_gas,temp)
  saveRDS(pge_20_gas, "pge_20_gas.rds")
}

year <- 2021
quarters <- 1:2
pge_21_gas <- NULL

for(quarter in quarters) {
  filename <- 
    paste0(
      "PGE_",
      year,
      "_Q",
      quarter,
      "_",
      type,
      "UsageByZip.csv"
    )
  print(filename)
  temp <- read_csv(filename)
  pge_21_gas <- rbind(pge_21_gas,temp)
  saveRDS(pge_21_gas, "pge_21_gas.rds")
}

pge_17_21_gas <- do.call("rbind", list(pge_17_gas, pge_18_gas, pge_19_gas, pge_20_gas, pge_21_gas))

saveRDS(pge_17_21_gas, "pge_17_21_gas.rds")
```

```{r merge datafiles}
pge_final <- merge (x=pge_17_21_elec, y=pge_17_21_gas, all=TRUE)
pge_filter_res <- filter(pge_final, CUSTOMERCLASS %in% c("Elec- Residential", "Gas- Residential"))
pge_filter_com <- filter(pge_final, CUSTOMERCLASS %in% c("Elec- Commercial", "Gas- Commercial"))
```

```{r data preprocessing}
pge_res_final <-
  pge_filter_res %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_res_final2 <-
  pge_filter_res %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_com_final <-
  pge_filter_com %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, MONTH, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU; no conversion needed for gas as 1therm = 1kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    DATE = 
      paste(
        YEAR,
        MONTH, 
        "01",
        sep="-"
      ) %>% as.Date()
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )

pge_com_final2 <-
  pge_filter_com %>%
  select(
    -c(COMBINED, AVERAGEKWH, AVERAGETHM)
  ) %>%
  group_by(YEAR, CUSTOMERCLASS) %>%
  summarize(
    TOTALKWH = 
      sum(
        TOTALKWH, na.rm = T
      ),
    TOTALTHM =
      sum(
        TOTALTHM, na.rm = T
      )
  ) %>%
  mutate(
    ELEC_KBTU = TOTALKWH*3.41 #1kWh=3.41kBTU; no conversion needed for gas as 1therm = 1kBTU
  ) %>%
  mutate(GAS_KBTU = TOTALTHM*100 #1therm=100kBTU
  ) %>%
  mutate(
    TOTALBTU = GAS_KBTU + ELEC_KBTU
  )
```

# PG&E Monthly Electricity and Gas Usage (Residential)

Both electricity and gas usage fluctuates seasonally, with electricity usage peaking over summer at about 1 x 10^10^ kBTU and gas usage peaking each winter at about 3.5 to 4.0 x 10^10^ kBTU. Generally, fluctuations in total energy usage is driven by gas usage due to the marked increase of more than 2 x 10^10^ kBTU during winter, compared to less than 0.5 x 10^10^ kBTU from electricity usage over summer. This can be explained by the role of heating in building systems during winter.

There is a less steep decline in gas usage since the start of the COVID-19 pandemic, with gas usage about 0.2 to 0.3 x 10^10^ kBTU higher in April 2020 compared to previous years.There is a new peak in electricity usage in residential buildings in August 2020 at 1.14 x 10^10^ kBTU due to people spending more time in their homes for work and education during the pandemic.

```{r electricity and gas residential chart - date}
library(plotly)

pge_chart_res1 <- 
  pge_res_final %>%
  ggplot() +
  geom_line(
    aes(
      x = DATE %>% factor(),
      y = TOTALBTU,
      color = CUSTOMERCLASS,
      group = 1
    ),
    stat = "identity",
    position = "identity"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Residential) from 2017 to 2021",
    color = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10)
  )

pge_chart_res1 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)
```

A key issue about analysing large datasets (i.e. 54 months across many zipcodes) is that it becomes hard to observe cumulative trends, e.g. by year rather than month. Grouping the data of residential sector into years (2021 data only includes the first two quarters), there is a clear increase in total energy usage between 2019 and 2020 during the COVID-19 pandemic. The interactive plot allows us to identify a bigger jump in electricity than gas usage (in absolute terms). Perhaps gas usage is less dynamic than electricity usage as heating is likely the main source and is less variable, compared to the diverse range of household and electrical applicances that require electricity when people are at home.

```{r electricity and gas residential chart - year}
pge_chart_res2 <- 
  pge_res_final2 %>%
  ggplot() +
  geom_bar(
    aes(
      x = YEAR %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Residential) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_res2 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)

```

# PG&E Monthly Electricity and Gas Usage (Commercial)

On the other hand, total electricity and gas usage in the commercial sector share similar seasonal fluctuations with the residential sector due to energy consumption by building heating systems. There is a sharp decrease in both electricity and gas use between January and June 2020, with electricity and gas use decreasing to about 6.4 x 10^9^ kBTU and 3.7 x 10^9^ kBTU respectively when the first wave of the COVID-19 pandemic occurred, which can be attributed to the introduction of remote working to reduce social interaction. 

There is another sharp drop in electricity usage during the second wave of COVID-19 in November and December 2020, where more people worked away from offices. This is consistent with a lower peak in gas usage in the same period due to a lesser need for building heating, bucking the trend of increasing gas usage during winter from 2017/18 to 2019/20. However, more recent data suggests that people are returning to offices, with electricity usage in the first half of 2021 reaching new peaks of 1.07 x 10^10^ kBTU in June 2021.

```{r electricity and gas commercial chart}
pge_chart_com1 <- 
  pge_com_final %>%
  ggplot() +
  geom_line(
    aes(
      x = DATE %>% factor(),
      y = TOTALBTU,
      color = CUSTOMERCLASS,
      group = 1
    ),
    stat = "identity",
    position = "identity"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Commercial) from 2017 to 2021",
    color = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_com1 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)
```

Likewise, grouping the data of the commercial sector into years (2021 data only includes the first two quarters), total energy consumption in 2020 was lower than that in 2019. This is expected due to more people working outside offices. However, the data also shows fluctuations in total energy use from 2017 to 2020, rather than an upward trend that was 'broken' by COVID-19. It must be acknowledged that other factors beyond the COVID-19 pandemic influencing energy use are also at play, such as cost of energy. While we may anticipate an increasing trend of energy consumption over time due to affluence, it may also be mitigated by better design of urban systems to improve energy efficiency.  

```{r electricity and gas commercial chart - year}
pge_chart_com2 <- 
  pge_com_final2 %>%
  ggplot() +
  geom_bar(
    aes(
      x = YEAR %>% factor(),
      y = TOTALBTU,
      fill = CUSTOMERCLASS
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Time",
    y = "kBTU",
    title = "PG&E Territory Monthly Electricity and Gas Usage (Commercial) from 2017 to 2021",
    fill = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(size=10))

pge_chart_com2 %>% 
  ggplotly() %>%
  layout(
    xaxis = list(fixedrange = T),
    yaxis = list(fixedrange = T)
  ) %>% 
  config(displayModeBar = F)
```


